The core transcriptome of mammalian placentas and the divergence of expression with placental shape

### 2017-02-06 manuscript/manuscript.Rnw
switch from 1:1 to single-copy ortholog

1:1 orthologs in were determined using
Ensembl v.80 gene trees via Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
in at least one species.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on 1:1 orthologs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$. For \human and \mouse lineage specific calculations, a
type-III anova was calculated using
lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
species having separate random effects. Shifts in expression levels
are reported with $\log_2$ Fold-Change (FC), rounded to the nearest
integer.

@@

print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual("Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      # scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      theme(text=element_text(size=9),
            legend.position=c(0.99,0.41),
            axis.text.y=element_text(face="italic",size=7),
            legend.margin=unit(0,"mm"),
            legend.justification=c(1,1),
            plot.margin=unit(c(0,0,0,0),"mm")
            ))
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all 13 eutherians studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom) by median expression. If we also require
  expression (\ac{FPKM}$ \ge 10$) in the marsupial, \opossum, there are 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 &
    `Min. Exp. in` == '14 of 14 Species',max(num_genes)]} genes
  which meet this criteria.
  \Cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different \ac{FPKM} and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

  shows the number of 1:1 orthologous genes included in the core

@@

  genes would be included if the expression of a set of 1:1 orthologs of all 13 Eutherians must be at

@@

  1:1 orthologs would be included if 13 of 14 species have an

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a 1:1 ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14 Species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express 1:1
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}). Supplemental
\cref{tab:core_placenta_pathway_analysis} shows the pathways with
significant and suggestive enrichment in the core placenta
transcriptome.

>>>>

Single-copy (1-to-1) orthologs in were determined using
Ensembl v.80 gene trees via Biomart using the biomaRt R package
\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
  1,length(unique(human_name))]} single-copy orthologous genes were expressed
in at least one species.

Significant shifts in expression levels between groups of species were
calculated using Student's t test on single-copy orthologs using the mean of
each species to avoid large biases towards samples with $n \gg 1$ and
considered significant if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
$\le 0.05$. For \human and \mouse lineage specific calculations, a
type-III anova was calculated using
lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
species having separate random effects. Shifts in expression levels
are reported with $\log_2$ Fold-Change (FC), rounded to the nearest
integer.

@@

print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
      geom_boxplot(outlier.shape=NA)+
      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
      geom_point(aes(color=Species,shape=Species))+
      scale_shape_manual("Species",
                         guide=guide_legend(#nrow=1,
                                            keywidth=unit(3,"mm"),
                                            keyheight=unit(3,"mm"),
                                            label.theme=element_text(face="italic",
                                                                     angle=0,
                                                                     size=7),
                                            title.position="top"
                                            ),
                         values=0:18)+
      # scale_color_discrete(NULL)+
      coord_flip()+
      xlab("Human Ortholog Name")+
      ylab("Gene Expression (FPKM)")+
      theme_bw() +
      theme(text=element_text(size=9),
            legend.position=c(0.99,0.41),
            axis.text.y=element_text(face="italic",size=7),
            legend.margin=unit(0,"mm"),
            legend.justification=c(1,1),
            plot.margin=unit(c(0,0,0,0),"mm")
            ))
@ 
\caption{Core Placenta Transcriptome.
  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
  single-copy \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
  placenta of all 13 eutherians studied with housekeeping genes
  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
  ordered (from top to bottom) by median expression. If we also require
  expression (\ac{FPKM}$ \ge 10$) in the marsupial, \opossum, there are 
  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 &
    `Min. Exp. in` == '14 of 14 Species',max(num_genes)]} genes
  which meet this criteria.
  \Cref{fig:core_placenta_transcriptome_shape} shows the shape of the
  distribution of the number of genes in the core placenta
  transcriptome if different \ac{FPKM} and species restriction criteria are
  utilized.}
\label{fig:core_placenta_transcriptome}
\end{figure}

@@

  shows the number of single-copy orthologous genes included in the core

@@

  genes would be included if the expression of a set of single-copy orthologs of all 13 Eutherians must be at

@@

  single-copy orthologs would be included if 13 of 14 species have an

@@

The
\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
genes comprising the core placenta transcriptome are shown in
\cref{fig:core_placenta_transcriptome} ordered by their median
transcription level in \ac{FPKM}. We considered a gene to be core to
the placenta transcriptome if a single-copy ortholog of that gene was
expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
(excluding the metatherian, \opossum) and was not a human housekeeping
gene according to
\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
meet this criteria but are housekeeping genes are depicted in
Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
which would be included in the core placenta set if the \ac{FPKM}
cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
of 1 yields %
\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
  == '14 of 14 Species',max(num_genes)]} %
genes in the core transcriptome. %
In contrast to the shape of the number of genes included in the core
placenta transcriptome, the set of \human housekeeping genes expressed
all placentas is flatter (Supplemental
\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
of the total fraction of genes. However, there is still a large
difference between the set of housekeeping genes which are expressed
in \human and those which are expressed in all placentas.

Supplemental \cref{tab:core_placenta_go_terms} lists the significant
enrichment of \ac{GO} terms in the core placenta transcriptome;
notable terms include those involved in cytoskeletal structure,
protein transport, membrane protein structure, and inflammation
resolution. Increasing the permissiveness of the core placenta
transcriptome by only requiring 10 of 14 species to express single-copy
orthologs with FPKM $>10$ and results in the same terms also being
significantly enriched, along with additional terms (Supplemental
\cref{tab:core_placenta_go_terms}). Supplemental
\cref{tab:core_placenta_pathway_analysis} shows the pathways with
significant and suggestive enrichment in the core placenta
transcriptome.
